Refinement for open automata

### 2021-05-10 2021-Refinement/report.tex
Proof for Lemma 2

The composition  \(A_p\subst{A_c}{k} = \OA{S}{s_0}{V}{\sigma_0}{J}{T}\) with \(A_c \defobject \OA{S_c}{s_{0c}}{V_c}{\sigma_{0c}}{J_c}{T_c}\), \(A_p \defobject \OA{S_p}{s_{0p}}{V_p}{\sigma_{0p}}{J_p}{T_p}\) and \(k \in J_p\), is a non-locking composition if there is a reachability predicate such that from reachable configurations, either a transition is possible in \(A_p\subst{A_c}{k}\) or there were no possible transition in \(A_p\) to begin with
\begin{multline*}
	\forall s \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{k}}\mpar{s}} \implies \mpar{\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}} \\
	\vee \forall t_p \in \fOT{\pi_{S_p}\mpar{s}}, \forall \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \nvdash \fguard{t_p}
\end{multline*}
\end{defi}

To illustrate the introduction of a deadlock in a composition and the definition of non-locking composition, we consider the following example.
\begin{exi}[Deadlocks introduced by composition]
% TODO: give a context, a circuit cannot block physical time from progressing
\Ludo{TODOludo}{lire l explication quand elle y sera}
\begin{figure}
\input{randomtick_clock.tex}
\vrule
\input{register_anytick.tex}
\caption{On the left: A clock which imposes a tick; On the right: A modified version of the counter at figure \ref{fig:tlh}}
\label{fig:anytick}
\end{figure}
% TODO: explain
\begin{figure}
\input{Composition_deadlock.tex}
\vrule
\input{Composition_nolock.tex}
\caption{On the left: Deadlocks introduced by composition; On the right: No deadlock introduced by composition}
\label{fig:deadlock}
\end{figure}
% TODO: explain
\end{exi}

@@

\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{k}\) is a non-locking composition if:
\[ \forall s \in S, \forall \sigma: V \uplus \bigcup_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
Proof of both lemma are given in appendix \ref{apx:lemeqd}.

@@

This new substitution can replace the standard substitution in the case where only one hole is filled and also in the parallel case.
In the parallel case the condition is checked after the parallel substitution because there is no guarantee that any order of sequential composition is non-locking for every substitution even if the parallel substitution is non-locking.
\begin{exi} % TODO
\begin{figure}
\centering
\input{}
\vrule
\input{}
\vrule
\input{}
\caption{}
\label{fig:pnls}
\end{figure}
An example of non-locking parallel substitution where no individual substitution is non-locking is given in Figure \ref{fig:pnls}.% TODO HERE
\end{exi}
From this point and in the previous definitions, composition will only refer to non-locking composition.

@@

\item[\(\impliedby\):] We admit the bottom formula as \(H_{ref}\), let \(t_1 \defobject \OTx{1}{}{1}{} \in T_1\) % TODO HERE; quantify on all valuations to get t_{2\sigma}

@@

\begin{proof}[Lemma 1]

@@

	\sigma \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \tag{WD}\label{eq:WD}

@@

	\vee \forall t_2 \in \fOT{s_2}, \forall \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \nvdash \fguard{t_2} \tag{ID}\label{eq:ID}

@@

\item[\bf(\ref{eq:WD})\(\implies\)(\ref{eq:ID}):]

@@

	By excluded middle, either the right hand side of the disjunction in (\ref{eq:WD}) holds, in which case the property holds, or it is false and we have the hypothesis:

@@

	\(R\) satisfies (\ref{eq:WD}) with the current value of \(\mpar{s_1, s_2}\) and \(\sigma \uplus \nu\) completed with dummy values for the other variables of \(\bigcup_{t_2 \in \fOT{s_2}} \fguard{t_2}\) so we have:
	\[ \sigma \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \hyp{3} \]

@@

\item[\bf(\ref{eq:WD})\(\impliedby\)(\ref{eq:ID}):]

@@

	To do that we use (\ref{eq:ID}) with the current value of \(\mpar{s_1, s_2}\) and \(\sigma\), and prove the premisse of the implication with \hyp{1} to get:

@@

\begin{proof}[Lemma 2]
% TODO: recall what we want to prove, then prove it
\end{proof}

>>>>

The composition  \(A_p\subst{A_c}{k} = \OA{S}{s_0}{V}{\sigma_0}{J}{T}\) with \(A_c \defobject \OA{S_c}{s_{0c}}{V_c}{\sigma_{0c}}{J_c}{T_c}\), \(A_p \defobject \OA{S_p}{s_{0p}}{V_p}{\sigma_{0p}}{J_p}{T_p}\) and \(k \in J_p\), is a non-locking composition if there is a reachability predicate such that from reachable configurations, either a transition is possible in \(A_p\subst{A_c}{k}\) or there were no possible transition in \(A_p\) to begin with
\begin{multline*}
	\forall s \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{k}}\mpar{s}} \implies \mpar{\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}} \\
	\vee \forall t_p \in \fOT{\pi_{S_p}\mpar{s}}, \forall \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \nvdash \fguard{t_p}
\end{multline*}
\end{defi}
(\(\pi_X\) is the projection of tuples on \(X\)).

To illustrate the introduction of a deadlock in a composition and the definition of non-locking composition, we consider the following example.
\begin{exi}[Deadlocks introduced by composition]
\Ludo{TODOludo}{lire l'explication}
This example reuses the counter introduced in Figure \ref{fig:tlh}.
The behaviour of the counter is to count external \(\act{tick}\) actions up to the amount set by action \(\act{set}\mpar{x}\) then to notify the environnement that this amount of time elapsed.
In the traffic light example the clock is not specified, we will introduce deadlocks using a wisely choosen clock specification.
However the full traffic light will not be used in order to simplify the example, only the counter will be.

\begin{figure}
\input{randomtick_clock.tex}
\vrule
\input{register_anytick.tex}
\caption{On the left: A clock which imposes a tick; On the right: A modified version of the counter at figure \ref{fig:tlh}}
\label{fig:anytick}
\end{figure}
The clock on the left side of Figure \ref{fig:anytick} transmit the actions of its hole unchanged until the clock imposes a \(\act{tick}\) on its hole.
This can model a physical clock, because physical time ticks cannot be delayed.

If the hole cannot register a \(\act{tick}\) at any time then there is a deadlock, which is the case with our counter.
The composition of this clock with the counter is given on the left of Figure \ref{fig:deadlock}.
\begin{figure}
\input{Composition_deadlock.tex}
\vrule
\input{Composition_nolock.tex}
\caption{On the left: Deadlocks introduced by composition; On the right: No deadlock introduced by composition}
\label{fig:deadlock}
\end{figure}
There are two deadlocks:
The first is in the state \(1S\), which correspond to the clock trying to imppose a tick when the counter is not set.
The second is in the state \(1C\), which correspond to the clock trying to impose a tick when the counter has reached the amount peviously set, but not has not yet reported it to the environment.
These deadlocks could have been in the specification in which case they were intended, but it is not the case here.

The counter on the right of Figure \ref{fig:anytick} is a modification to accept ticks at in any state.
Composing this modified counter gives the automaton on the right of Figure \ref{fig:deadlock} which has no deadlocks.

We should successfully caracterise the second composition as non-locking but fail for the first.
In the state \(0S\) any valuation is reachable (no initial valuation).
In both automata the transition \nmm{\OT{0S}{1S}{\tau}{\mbrc{}}{\top}{\mbrc{}}} has a true guard, hence the first branch of the disjunction holds.
In the state \(0C\) % TODO: run the definition of non-locking composition
\end{exi}

@@

\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{k}\) is a non-locking composition if:
\[ \forall s \in S, \forall \sigma: V \uplus \bigcup_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
Proof of both lemma are given in appendix \ref{apx:lemeqd}.

@@

This new substitution can replace the standard substitution in the case where only one hole is filled and also in the parallel case.
In the parallel case the condition is checked after the parallel substitution because there is no guarantee that any order of sequential composition is non-locking for every substitution even if the parallel substitution is non-locking.
\begin{exi} % TODO
\begin{figure}
\centering
\input{}
\vrule
\input{}
\vrule
\input{}
\caption{}
\label{fig:pnls}
\end{figure}
An example of non-locking parallel substitution where no individual substitution is non-locking is given in Figure \ref{fig:pnls}.% TODO
\end{exi}
From this point and in the previous definitions, composition will only refer to non-locking composition.

@@

\item[\(\impliedby\):] We admit the bottom formula as \(H_{ref}\), let \(t_1 \defobject \OTx{1}{}{1}{} \in T_1\) % TODO; quantify on all valuations to get t_{2\sigma}

@@

\begin{proof}[Proof of lemma 1]

@@

	\sigma \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \tag{WD}\label{eq:drWD}

@@

	\vee \forall t_2 \in \fOT{s_2}, \forall \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \nvdash \fguard{t_2} \tag{ID}\label{eq:drID}

@@

\item[\bf(\ref{eq:drWD})\(\implies\)(\ref{eq:drID}):]

@@

	By excluded middle, either the right hand side of the disjunction in (\ref{eq:drWD}) holds, in which case the property holds, or it doesn't hold and we have the hypothesis:

@@

	\(R\) satisfies (\ref{eq:drWD}) with the current value of \(\mpar{s_1, s_2}\) and \(\sigma \uplus \nu\) completed with dummy values for the other variables of \(\bigcup_{t_2 \in \fOT{s_2}} \fguard{t_2}\) so we have:
	\[ \sigma \uplus \nu \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \hyp{3} \]

@@

\item[\bf(\ref{eq:drWD})\(\impliedby\)(\ref{eq:drID}):]

@@

	To do that we use (\ref{eq:drID}) with the current value of \(\mpar{s_1, s_2}\) and \(\sigma\), and prove the premisse of the implication with \hyp{1} to get:

@@

\begin{proof}[Proof of lemma 2]
\[ \forall s \in S, \forall \sigma: V \uplus \bigcup_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \tag{WD}\label{eq:nlcWD} \]
\[ \iff \]
\begin{multline*}
	\forall s \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A}\mpar{s}} \implies \mpar{\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}} \\
	\vee \forall t_p \in \fOT{\pi_{S_p}\mpar{s}}, \forall \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \nvdash \fguard{t_p} \tag{ID}\label{eq:nlcID}
\end{multline*}
\item[\bf(\ref{eq:nlcWD})\(\implies\)(\ref{eq:nlcID}):]
	Let \(s \in S\) and \(\sigma \in V \to \values\) be such that:
	\[ \sigma \vdash \reach{A}\mpar{s} \hyp{1} \]
	By excluded middle, either the right hand side of the disjunction in (\ref{eq:nlcWD}) holds, in which case the property holds, or it doesn't hold and we have the hypothesis:
	\[ \exists t_p \in \fOT{\pi_{S_p}\mpar{s}}, \exists \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \vdash \fguard{t_p} \hyp{2} \]
	We get immediately the values \(t_p\), \(\nu\) and the hypothesis:
	\[ \sigma \uplus \nu \vdash \fguard{t_p} \hyp{2'} \]
	\(R\) satisfies (\ref{eq:nlcWD}) with the current value of \(s\) and \(\sigma \uplus \nu\) completed with dummy values for the other variables of \(\bigcup_{t_p \in \fOT{\pi_{s_p}\mpar{s}}} \fguard{t_p}\) so we have:
	\[ \sigma \uplus \nu \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \hyp{3} \]
	The \(\vdash\) hides an implicit \(\exists\) that we use to get values to the variables in \(\bigcup_{t \in \fOT{s}}\) in a valuation \(\mu\).
	The left side of the implication is proved using \hyp{1} and \hyp{2'} for the branch \(t_p\) of the disjunction.
	We are now allowed to decompose the right side to get \(t\) such that:
	\[ \sigma \uplus \nu \uplus \mu \vdash \fguard{t} \hyp{3'} \]
	At this point we have a \(t\) which is composed of a transition in \(A_p\) and one in \(A_c\) so unlike lemma 1 we do not need to check that two transitions matches.
	The current goal is:
	\[ \exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t} \]
	Which is proved on \(t\) with valuation \(\mu\) by \hyp{3'}.
\item[\bf(\ref{eq:nlcWD})\(\impliedby\)(\ref{eq:nlcID}):]
	Let \(s \in S\), \(\sigma \in V \uplus \bigcup_{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fvars{t_p} \to \values\).
	We want to prove:
	\[ \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{\pi_{S_p}\mpar{s}}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \goal{1} \]
	We can admit the left part of the implication in which we decompose the big disjunction to get \(t_p\) and:
	\begin{align*}
		\sigma \vdash \reach{A}\mpar{s} & \text{ \hyp{1}} &
		\sigma \vdash \fguard{t_p} & \text{ \hyp{2}}
	\end{align*}
	We are left to prove:
	\[ \sigma \vdash \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \goal{1'} \]
	To do that we use (\ref{eq:nlcID}) with the current value of \(s\) and \(\sigma\), and prove the premisse of the implication with \hyp{1} to get:
	\begin{multline}
		\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t} \\
		\vee \forall t_p \in \fOT{\pi_{S_p}\mpar{s}}, \forall \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \nvdash \fguard{t_p} \hyp{3}
	\end{multline}
	There are two cases in this disjunction, the second is in contradiction with \hyp{2} and the first gives \(t\) and \(\nu\) such that:
	\[ \sigma \uplus \nu \vdash \fguard{t} \hyp{4} \]
	Which proves \goal{1'} with the values of \(\nu\) for the variables of \(\fguard{t}\) and dummy values for the other guards.
\end{proof}
